---
title: "Project Markdown: Sports Bet like a Hedge Fund Manager"
author: "Nico Manzonelli, Jeffrey Mayolo, Joseph Zuccarelli, Simon Lam"
date:
output: pdf_document
---

## Load Packages

```{r,eval=T, warning=F, message=F, results='hide'}
# Import necessary libraries
library(dplyr)
library(ggplot2)
library(tidyr)
library(C50)
library(randomForest)
```

## Read in the data

```{r eval=T, message=FALSE, warning=FALSE}
#Read in the data and do some preliminary cleaning to merge the two df
nfl <- read.csv('nfl_odds.csv')
ncaa <- read.csv('ncaa_odds.csv')

ncaa <- ncaa %>%
  select(-home_plays, -away_plays) %>%
  na.omit()
ncaa <- ncaa[,2:length(ncaa)]

nfl['nfl'] <- 1
ncaa['nfl'] <- 0

all_games <- rbind(ncaa, select(nfl, -X))

#Extract the median odds for all bookmakers
# Make a column for underdog
# Make a column for home money multiplier at start and half
# Make column for change in money multiplier
# Impute NAs with mean
# Double check there are no more NAs
medians <- all_games %>%
  select(-bookmaker) %>%
  group_by(home_team, away_team, date) %>%
  summarise_all(median, na.rm = T) %>%
  mutate(home_underdog = ifelse(home_odds_start > away_odds_start, 1, 0)) %>%
  mutate(home_mul_start = ifelse(home_odds_start > 0, home_odds_start / 100, 100 / abs(home_odds_start))) %>%
  mutate(home_mul_half = ifelse(home_odds_half > 0, home_odds_half / 100, 100 / abs(home_odds_half))) %>%
  mutate(change_in_mul = home_mul_start - home_mul_half) %>%
  mutate(point_diff_half = home_score_half - away_score_half) %>%
  mutate(home_pen = ifelse(is.na(home_pen), mean(all_games$home_pen, na.rm = T), home_pen)) %>%
  mutate(away_pen = ifelse(is.na(away_pen), mean(all_games$away_pen, na.rm = T), away_pen)) %>%
  na.omit()


shoped_odds <- all_games %>% 
  select(-bookmaker) %>%
  group_by(home_team, away_team, date) %>%
  summarise_all(max, na.rm = T) %>%
  mutate(home_underdog = ifelse(home_odds_start > away_odds_start, 1, 0)) %>%
  mutate(home_mul_start = ifelse(home_odds_start > 0, home_odds_start / 100, 100 / abs(home_odds_start))) %>%
  mutate(home_mul_half = ifelse(home_odds_half > 0, home_odds_half / 100, 100 / abs(home_odds_half))) %>%
  mutate(change_in_mul = home_mul_start - home_mul_half) %>%
  mutate(point_diff_half = home_score_half - away_score_half) %>%
  mutate(home_pen = ifelse(is.infinite(home_pen), mean(all_games$home_pen, na.rm = T), home_pen)) %>%
  mutate(away_pen = ifelse(is.infinite(away_pen), mean(all_games$away_pen, na.rm = T), away_pen)) %>%
  na.omit()


```
#EDA

```{r, eval=T, warning=F, message=F}
# Histogram: Change in Home Team Odds (Initial - Halftime)
medians %>%
  ggplot(aes(change_in_mul)) +
  geom_histogram(bins = 30) +
  labs(x = "Home Team Change in MM (Start - Half)", y = "Count", 
       title = "Histogram: Home Team Change in Money Multiplier (Start - Half)") +
  theme_minimal()
```

Finally, let's explore the relationship between the response variable, the home team's change in odds at halftime, and one of our predictors, the score differential at halftime (home score - away score). Notice that there appears to be a positive linear association between the home team's change in odds at halftime and the score differential at halftime.

```{r eval=F, message=FALSE, warning=FALSE}
# Plot: Home Team Change in Odds (Start-Half) vs. Halftime Score Differential
medians %>%
  ggplot(aes(x=point_diff_half, y=change_in_mul, color = as.factor(nfl))) + 
  geom_point() + 
  geom_smooth(method = 'lm') +
  xlab("Halftime Score Differential (Home - Away)") +
  ylab("Halftime Change in Money Multiplier (Start - Half)") +
  ggtitle("Home Team Change in Money Multiplier vs. Score Differential (Halftime)") +
  labs(color = "Game") +
  scale_color_discrete(c("NCAAF", "NFL")) +
  theme_minimal()
  
```

# Preliminary Model

```{r,eval=T}
# Base Model (home_odds_change ~ score_differential_half)
baseModel <- lm(change_in_mul ~ point_diff_half * home_underdog + nfl*point_diff_half, data = medians)
summary(baseModel)
plot(baseModel, which = c(1,2))
```
```{r,eval=T}
# Weighted Model (home_odds_change ~ score_differential_half)
weightedModel <- lm(change_in_mul ~ point_diff_half * home_underdog + nfl*point_diff_half, 
                weights = point_diff_half^2, data = medians)
summary(weightedModel)
plot(weightedModel, which = c(1,2))
```

```{r,eval=T}
# Ploy Model (home_odds_change ~ score_differential_half)
ployModel <- lm(change_in_mul ~ poly(point_diff_half, 2, raw = T), data = medians)
summary(ployModel)
plot(ployModel, which = c(1,2))
```

## Confounders? Multi-Regression

```{r,eval=T, out.height=.5}
#make every game statistic a difference term
multi <- medians %>%
  ungroup() %>%
  mutate(yards_diff = home_yards - away_yards) %>%
  mutate(first_down_diff = home_first_downs - away_first_downs) %>%
  mutate(fumbles_diff = home_fumbles - away_fumbles) %>%
  mutate(sacks_diff = home_sacks - away_sacks) %>%
  mutate(ints_diff = home_ints - away_ints) %>%
  mutate(penalty_diff = home_pen - away_pen)

#define the multi model with poly
multiModel <- lm(change_in_mul ~ (poly(point_diff_half, 2, raw = T) + yards_diff + first_down_diff + fumbles_diff + sacks_diff + ints_diff + penalty_diff), data = multi)

summary(multiModel)
plot(multiModel, which = c(1,2))
```

## Prediction Modeling

Train-Test Split

```{r}
###SPLIT TEST AND TRAIN#########
set.seed(139)
train_idx <- sample(1:nrow(medians), 88)
train <- medians[train_idx,]
test <- medians[-train_idx,]

X_train <- train %>%
  ungroup() %>%
  select(point_diff_half , home_yards, away_yards,
         home_first_downs, away_first_downs, home_fumbles, away_fumbles,
         home_sacks, away_sacks, home_ints, away_ints, home_pen, away_pen, 
         espn_win_pred, home_underdog, nfl)
y_train <- as.factor(train$home_win)

X_test <- test %>%
  ungroup() %>%
  select(point_diff_half , home_yards, away_yards,
         home_first_downs, away_first_downs, home_fumbles, away_fumbles,
         home_sacks, away_sacks, home_ints, away_ints, home_pen, away_pen, 
         espn_win_pred, home_underdog, nfl)
y_test <- as.factor(test$home_win)
```

Fit a single tree

```{r}
######FIT A SINGLE DECISION TREE########
tree <- C5.0(x = X_train, y = y_train)

plot(tree)

#training accuracy
sum(predict(tree, X_train) == y_train) / nrow(X_train)
#testing accuracy
sum(predict(tree, X_test) == y_test) / nrow(X_test)

```

Fit a Random Forest

```{r}
####FIT A WELL TUNED RANDOM FOREST######
grid = expand.grid(maxnodes = c(6,12,25,50,75,100), mtrys = c(4, 6, 12,16))
val_scores <- rep(NA, nrow(grid))
X_train["home_win"] = y_train

#cross validate
for (i in 1:nrow(grid)) {
  rf <- randomForest(home_win ~ ., data = X_train, 
                     maxnodes= grid[i,1], mtry = grid[i,2], ntree = 300)
  val_scores[i] <- sum(rf$predicted == X_train$home_win, na.rm = T) / length(na.omit(rf$predicted))
}

#extract best hyperparamters for tuning
best_grid <- which(val_scores == min(val_scores))

#recreate best model
rf <- randomForest(home_win ~., data = X_train,
                    maxnodes= grid[best_grid, 1], mtry = grid[best_grid, 2], ntree = 300)

#extract variable importance
variable_importance <- data.frame(MeanDecreaseGini = rf$importance[], Variable = names(rf$importance[,1]))

ggplot(variable_importance) +
  geom_bar(aes(x = reorder(Variable, MeanDecreaseGini), y = MeanDecreaseGini), 
           stat = 'identity', alpha = .8) +
  coord_flip() +
  ylab("Mean Decrease In Gini Index At Each Split") +
  xlab("Variable") +
  ggtitle("Random Forest Variable Importance") +
  theme_minimal()


#training accuracy
sum(predict(rf, X_train) == y_train) / nrow(X_train)

#testing accuracy
sum(predict(rf, X_test) == y_test) / nrow(X_test)
```


## Profit Simulatuion

```{r}
#Imagine 1000 different bettors betting on the same week. 
#This simulation gets the difference in money made for each better if they followed our
#strategy or threw or decided not to hedge

simulate_profit_difference <- function(sim_test, model) {
  
  sim_test["model_prediction"] <- predict(model, sim_test)
  profit_diff <- c()
  
  for (i in 1:1000) {
  informed_stuff <- sim_test %>%
    mutate(underdog_win = ifelse(home_underdog == home_win, 1, 0)) %>%
    mutate(bet_underdog = sample(0:1, 1, prob = c(.5, .5))) %>%
    mutate(pred_underdog_win = ifelse(model_prediction == home_underdog, 1,0)) %>%
    mutate(hedge = ifelse(pred_underdog_win == bet_underdog, 0, 1)) %>%
    mutate(init_odds = ifelse(home_underdog == bet_underdog, home_odds_start, away_odds_start)) %>%
    mutate(init_mul = ifelse(init_odds > 0, init_odds / 100, 100 / abs(init_odds))) %>%
    mutate(init_win = ifelse(bet_underdog == underdog_win, 1, 0)) %>%
    mutate(init_wins_prof = ifelse(init_win == 1, 100*init_mul, -100)) %>%
    mutate(hedge_odds_half = ifelse(home_underdog == 1, away_odds_half, home_odds_half)) %>%
    mutate(hedge_mul = ifelse(hedge_odds_half > 0, hedge_odds_half / 100, 100 / abs(hedge_odds_half))) %>%
    mutate(hedge_amount = (100*init_mul + 100) / (hedge_mul + 1)) %>%
    mutate(if_hedge_wins_prof = hedge_amount * hedge_mul) %>%
    mutate(if_hedge_prof = ifelse(init_win == 1, init_wins_prof - hedge_amount, if_hedge_wins_prof - 100)) %>%
    mutate(total_profit = ifelse(hedge == 1, if_hedge_prof, init_wins_prof))
  
    profit_diff[i] <- sum(informed_stuff$total_profit) - sum(informed_stuff$init_wins_prof)
  }
  return(profit_diff)
}

sim_test <- test 
tree_profit_diff <- simulate_profit_difference(sim_test, tree)
rf_profit_diff <- simulate_profit_difference(sim_test, rf)

profits <- rbind(data.frame(profit = tree_profit_diff, Model = "Single Tree"),
                 data.frame(profit = rf_profit_diff, Model = "Random Forest"))

ggplot(profits) +
  geom_histogram(aes(profit, fill = Model), alpha = .7) +
  ggtitle("Distribution of Difference in Profits") +
  xlab("Difference in Profits (Model Recommended Hedge - No Hedge)") +
  ylab("Frequency") +
  theme_minimal()


summary(tree_profit_diff)
summary(rf_profit_diff)
```